---
title: "Introduction to survAudit: Auditing Cox Proportional Hazards Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)
```

## Survival Analysis and the Cox Model

When conducting research where the outcome of interest is the time until an event occurs, the most widely used method is the **Cox Proportional Hazards (PH) model**. For a Cox model's estimates to be valid, the model must satisfy several structural and mathematical assumptions. 

Checking these assumptions usually requires running fragmented tests from multiple packages. `survAudit` solves this by providing:

1. **A Unified Assumption Classification**: Organizing assumptions into *Non-Identifiable* (purely qualitative), *Partially Assessable* (combining data and judgment), and *Statistically Assessable* (testable from data).
2. **Comprehensive Diagnostics**: It aggregates proportional hazards, functional form, influence, outliers, and generalized collinearity (GVIF) into one run.
3. **Auditable S3 Objects**: It returns an object containing all raw diagnostic metrics and lets analysts append qualitative justifications directly to the object before saving it.

---

## Case Study Workflow

In this vignette, we demonstrate how `survAudit` changes the analytical workflow of a typical survival analysis by revealing model limitations that are not apparent from standard Cox regression output. We use the classic `pbc` dataset from the `survival` package, which records survival data for patients with primary biliary cholangitis.

### 1. Fit the Initial Model

A standard approach might begin by fitting a multivariate Cox model using a mix of continuous covariates (`age`, `bili`, `albumin`, `protime`) and a categorical factor (`edema`):

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

# Load data and convert endpoint
data(pbc, package = "survival")
pbc$status <- ifelse(pbc$status == 2, 1, 0)

# Fit a naive Cox PH model
fit <- coxph(
  Surv(time, status) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
```

In many applied analyses, reporting may proceed at this stage, with interpretation focused primarily on coefficient estimates and significance tests.

### 2. Run the Initial Audit

However, `survAudit` goes beyond asking whether a model is significant. It systematically answers: *What aspects of this model require attention, and what can be learned from the available data versus what requires scientific justification?* 

We run the diagnostic audit using `survAudit()`. Supplying `data = pbc` is highly recommended to ensure complete case extraction for functional form plots.

```{r audit}
audit <- survAudit(fit, data = pbc)
summary(audit)
```

Running `survAudit()` immediately identifies several diagnostic findings that indicate areas where the model may require refinement. For example, the functional form diagnostic detects severe departures from linearity for both bilirubin and prothrombin time, while the proportional hazards test flags a global violation.

We can visually explore the functional form violations:

```{r plot-functional}
plot(audit, which = "functional")
```

Furthermore, the audit flags multiple highly influential observations and outliers. This explicitly prompts the analyst to check the raw data. Rather than automatically excluding these observations, `survAudit` preserves them as diagnostic findings requiring contextual evaluation.

### 3. Model Respecification: An Iterative Process

Crucially, the audit provides a transparent roadmap for model respecification. First, we applied clinically plausible logarithmic transformations to bilirubin and prothrombin time to address the detected departures from linearity.

```{r respecify}
# Fit an updated model with log transformations
fit2 <- coxph(
  Surv(time, status) ~ age + log(bili) + albumin + log(protime) + factor(edema),
  data = pbc
)

# Re-audit the updated model
audit2 <- survAudit(fit2, data = pbc)
summary(audit2)
```

A subsequent re-audit of the updated model reveals the impact of these changes. Importantly, the re-audit demonstrates that model correction is iterative rather than binary: while the functional form violations were substantially reduced, the remaining proportional hazards violation for `log(protime)` indicates that additional modeling decisions—such as introducing a time-dependent effect or stratification—may still be required.

### 4. The Auditable S3 Object

Importantly, `survAudit` does not just produce plots or terminal output; it returns a rich S3 object. This allows users to inspect individual diagnostic components, reuse computed quantities, and integrate audit results into custom reporting workflows.

For example, you can extract the underlying Proportional Hazards test table:

```{r extract}
print(audit2$ph$table)
```

Some assumptions (like Independent Censoring) cannot be tested statistically. `survAudit` requires you to document your qualitative justification. You can append these directly to the object:

```{r justify}
audit2$assumptions$non_identifiable$independent_censoring$justification <- 
  "Censoring was purely administrative (end of study), satisfying independent censoring."
```

If we print the audit again, this justification is now permanently attached to the object, ensuring complete transparency of both the statistics and your clinical reasoning.

By formalizing this iterative diagnostic process, `survAudit` reduces the likelihood that important model inadequacies remain undetected during routine analysis.
