---
title: "Introduction to survAudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5.5,
  fig.align = "center",
  dpi = 300,
  out.width = "100%",
  warning = FALSE
)
```

`survAudit` provides a structured diagnostic workflow for Cox proportional hazards models. It integrates proportional hazards testing, functional form assessment, influence and outlier diagnostics, collinearity checks, and event sufficiency into a single S3 object.

In routine practice, model diagnostics often either rely exclusively on p-values or ignore unverifiable assumptions altogether. A central design principle of `survAudit` is that model assumptions differ fundamentally in how far they can be evaluated from observed data alone:

- **Non-Identifiable**: Assumptions that cannot be verified statistically from data alone (e.g., independent censoring, absence of unmeasured confounding). These require substantive domain knowledge and study-design arguments.
- **Partially Assessable**: Assumptions informed by residual metrics but requiring visual inspection and analyst judgment (e.g., functional form, outlier impact).
- **Statistically Assessable**: Assumptions testable via formal statistical hypotheses (e.g., proportional hazards, collinearity, event sufficiency).

This tri-partite classification directly structures the audit outputs and reporting workflow illustrated below. The underlying residual diagnostics implement established methodology from the survival literature (Therneau & Grambsch, 2000; Klein & Moeschberger, 2003). The primary focus of `survAudit` is diagnostic detection and transparent documentation. While this vignette illustrates an initial respecification cycle, comprehensive discussions of remedial modeling strategies (such as splines, stratification, or time-varying coefficients) can be found in the broader survival analysis literature.

---

## Case Study: Primary Biliary Cholangitis

We use the `pbc` dataset from the `survival` package to illustrate a typical audit workflow.

### 1. Fit an Initial Model

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

data(pbc, package = "survival")

fit <- coxph(
  Surv(time, status == 2) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
summary(fit)
```

The model shows strong discrimination (Concordance = 0.813) and four predictors with $p < 0.001$. Standard Cox output contains no information about whether model assumptions are met, it is entirely focused on estimation.

### 2. Run the Audit

Pass the fitted model and the original dataset to `survAudit()`. The `data` argument is required because partial models must be refitted when assessing functional form.

```{r audit}
audit1 <- survAudit(fit, data = pbc)
audit1
```

The compact print output provides a quick orientation. The `[ ]` checkboxes for non-identifiable assumptions are intentionally unfilled: these cannot be evaluated from data and require the analyst to supply a design-based justification (see the last section). The statistically assessable block immediately flags proportional hazards violations for `bili` and `protime`, and reports an EPV of 26.7, which is adequate.

`summary()` provides the full diagnostic report including numeric results:

```{r audit-summary}
summary(audit1)
```

The detailed report adds the full Schoenfeld test table (with per-covariate chi-squared values), the five most influential observations by max |DFBETAs|, and the five largest deviance residuals, all with the original row identifiers from `pbc`. (Note: If you only need specific diagnostics programmatically without running the full report, each component is directly accessible, e.g., `audit1$ph$table` or `audit1$vif`; see [Accessing the Underlying Data](#accessing-the-underlying-data) below.)

Working through the output section by section:

- **Data context**: `protime` has 2 missing values, so the model is fitted on 416 of 418 observations. Ten event times are tied (6.2%), which is acceptable for the Efron approximation.
- **Non-identifiable assumptions**: Both independent censoring and absence of unmeasured confounding appear as unfilled checkboxes. They require explicit justification from study design; the package cannot evaluate them.
- **Goodness-of-fit**: Only signals that the Cox-Snell residuals were computed; the visual assessment is deferred to `plot(audit, which = "gof")`.
- **Functional form / Linearity**: Assesses whether continuous covariates have a linear relationship with the log hazard. Because there is no omnibus statistical test for non-linearity in Cox models, `survAudit` does not compute a p-value here. Instead, it lists the continuous predictors (`age`, `bili`, `albumin`, `protime`) for visual inspection via `plot(audit, which = "functional")`.
- **Influence**: 69 observations exceed the $2/\sqrt{n} = 0.098$ threshold. Observation 325 has the largest maximum |DFBETAs| of 0.96, which is sizeable relative to the threshold, indicating high sensitivity of at least one coefficient to this case.
- **Outliers**: 19 observations have |deviance| > 1.96 and 9 have |log-odds| > 3.66. The largest deviance residual is 3.24 (observation 319), suggesting this patient experienced the event substantially earlier than the model predicted.
- **Proportional hazards**: Evaluated using Schoenfeld residuals under the null hypothesis of constant hazard ratios over time ($H_0: \beta(t) = \beta$). A statistically significant test ($p < 0.05$) indicates an assumption violation. Importantly, a non-significant test does not rule out non-proportionality, as the test has limited power against non-monotone patterns or in smaller samples. Here, the global test is statistically significant ($p = 0.001$), with `protime` ($p = 0.003$) and `bili` ($p = 0.005$) individually flagged.

  Two limitations of the proportional hazards test are worth keeping in mind: (1) it detects only linear trends in the Schoenfeld residuals over time and can miss non-monotone violations; and (2) like any chi-squared test, it becomes increasingly sensitive to minor deviations as sample size grows. Visual inspection of the residual plots is therefore essential for assessing whether a statistically flagged violation is substantively meaningful.
- **EPV**: 160 events across 6 parameters gives an EPV of 26.7, which is adequate (threshold ≥ 20).
- **Collinearity**: All VIFs are near 1, indicating no meaningful collinearity.

The primary statistical issue is the proportional hazards violation, while the linearity of continuous predictors must be evaluated visually before deciding on model respecification.


### 3. Visual Diagnostics

#### Functional Form

```{r plot-functional}
plot(audit1, which = "functional")
```

For models with many continuous covariates, `plot()` arranges the panels into a multi-row grid. You can also inspect specific covariates individually using the `vars` argument (e.g., `plot(audit1, which = "functional", vars = "bili")` or `vars = c("bili", "protime")`).

For each continuous covariate, a restricted Cox model is fitted that omits that covariate. The martingale residuals from that restricted model are then plotted against the covariate values. Because the restricted model omits the covariate, its residuals will pick up the covariate's effect on the outcome. If the relationship is correctly specified as linear, the LOESS smooth should trace an approximately straight line. A curved or non-monotone smooth indicates that the linear specification is inadequate and that a transformation or more flexible term may be needed.

For `bili`, the smooth rises sharply at low values and flattens at high values, tracing a logarithmic curve consistent with a log-transformation. For `protime`, the smooth is approximately flat at low values before rising sharply and then dropping, again suggesting a non-linear relationship. In contrast, `age` and `albumin` track the linear reference lines closely, with the linear fit remaining well within the LOESS confidence bands throughout their range.

Common remedies for non-linearity include log-transforming the covariate, using polynomial or spline terms, or categorising the variable if a meaningful clinical threshold exists.

#### Proportional Hazards

```{r plot-ph}
plot(audit1, which = "ph")
```

The proportional hazards assumption requires that the log hazard ratio for each covariate is constant over time. The Schoenfeld residuals, after scaling, should be randomly scattered around a flat line if the assumption holds. A systematic slope or curve indicates that the hazard ratio changes over follow-up time.

The formal test (`cox.zph`) detects linear trends in the residuals over time. It will miss violations where the effect first increases and then decreases, or other non-monotone patterns. It is also sensitive to sample size: with several hundred observations, even a small and clinically irrelevant drift can yield a significant p-value. The plot is therefore the primary diagnostic; the p-value is supporting evidence.

For `bili` (p = 0.005) and `protime` (p = 0.003), the smoothed residual plots show clear trends over time. For `protime`, the hazard ratio appears to decrease over follow-up, meaning its prognostic effect is stronger early in follow-up than later. This is an important finding: a single time-averaged hazard ratio obscures this time-varying behaviour. For the remaining covariates, the smoothed curves remain roughly horizontal and easily encompass the reference line within their confidence bands.

#### Goodness of Fit

```{r plot-gof-initial}
plot(audit1, which = "gof")
```

The Cox-Snell residuals, if the model is correctly specified, should follow a unit exponential distribution. In the cumulative hazard plot, a well-fitting model will track the 45-degree reference line closely. Here the curve follows the diagonal through most of the range, indicating acceptable macro-level calibration. A modest deviation in the upper tail is common and typically reflects data sparsity at long follow-up times rather than systematic misfit. Crucially, goodness-of-fit is an omnibus, macro-level assessment. Even though the curve tracks the diagonal reasonably well overall, this global check is blind to the local non-linearities and proportional hazards violations identified above. An analyst who relies solely on global calibration would be falsely reassured. We therefore proceed to Section 4 to respecify the model.


### 4. Model Respecification

The functional form plots suggest log-transformations for `bili` and `protime`. We refit and re-audit:

```{r respecify}
fit2 <- coxph(
  Surv(time, status == 2) ~ age + log(bili) + albumin + log(protime) + factor(edema),
  data = pbc
)

audit2 <- survAudit(fit2, data = pbc)
summary(audit2)
```

Comparing this report to `audit1` highlights what the transformations did and did not fix. On the positive side, the PH violation for `bili` is resolved (p = 0.306), and global collinearity remains negligible. On the negative side, the PH violation for `log(protime)` persists (p = 0.003), the global PH test is still significant (p = 0.027), and the number of flagged influential observations has increased slightly (74 vs. 69). The visual diagnostics below examine each of these in detail.

#### Functional Form Re-Assessment

```{r audit-corrected-functional}
plot(audit2, which = "functional")
```

The smooth for `log(bili)` is now approximately linear, confirming that the log scale captures the relationship adequately. The smooth for `log(protime)` still shows curvature, indicating that the log transformation did not fully resolve the non-linearity. Further exploration, such as a square-root transformation, splines, or categorisation, would be warranted before finalising the model.

#### Proportional Hazards Re-Assessment

```{r audit-corrected-ph}
plot(audit2, which = "ph")
```

The PH assumption is now satisfied for `log(bili)` (p = 0.306): the time-varying behaviour seen in the initial model was largely a consequence of the non-linear scale rather than a genuine change in the hazard ratio over time. The violation for `log(protime)` persists (p = 0.003), suggesting its effect does change over follow-up regardless of scale. Common approaches to address a genuine proportional hazards violation include stratifying the model on the offending variable or allowing its effect to vary over time by including an interaction between the covariate and a function of time (i.e., fitting a time-varying coefficient).

#### Influence

```{r influence-plot}
plot(audit2, which = "influence")
```

DFBETAs quantify how much each coefficient changes when an individual observation is removed. Values exceeding the size-adjusted $2/\sqrt{n}$ threshold are flagged. Because this cutoff shrinks with sample size, it naturally flags many observations in larger datasets and serves as a relative screen. Rather than inspecting every flagged point, analysts should focus on the extreme upper tail reported in the `summary()` table, where observations 325 and 293 show the largest values (full matrix in `audit2$influence$dfbetas`).

A high DFBETAs value is not by itself a reason to exclude a case. The appropriate response is to inspect extreme cases for data entry errors, assess whether they represent a distinct subgroup, and evaluate their impact through sensitivity analysis.

#### Outliers

```{r outliers-plot}
plot(audit2, which = "outliers")
```

Unlike influence diagnostics, which assess effects on model coefficients, outlier residuals identify individuals with unusual survival times given their covariate profile. Deviance residuals exceeding $\pm 1.96$ and log-odds residuals exceeding $\pm 3.66$ are flagged. Because roughly 5% of observations exceed $\pm 1.96$ by chance alone, a moderate number of flagged cases is expected in larger datasets. Attention should focus on extreme cases reported in the `summary()` table, such as observations 87 and 319 with deviance residuals above 3, where patients died substantially earlier than their prognosis suggested (full residuals in `audit2$outliers`).

As with influence, extreme cases should be examined individually to check for data quality issues, unrecorded events, or genuinely unusual clinical trajectories.

---

## Accessing the Underlying Data

All diagnostic results are stored in the returned S3 object and can be extracted directly. This is useful for custom reporting, sensitivity analyses, or integration into automated pipelines:

```{r extract}
# The full cox.zph table
audit2$ph$table

# Martingale residuals from the full model
head(audit2$outliers$martingale)

# Flagged influential observations
audit2$influence$flagged_obs
```

## Documenting Non-Identifiable Assumptions

For assumptions that cannot be tested statistically, `survAudit` provides the `document_assumption()` helper to record qualitative justifications directly on the audit object:

```{r justify}
audit2 <- document_assumption(
  audit2,
  assumption = "independent_censoring",
  justification = "Censoring was purely administrative (end-of-study cutoff). No competing events or loss to follow-up related to prognosis."
)

# The recorded text is stored directly within the audit object:
audit2$assumptions$non_identifiable$independent_censoring$justification
```

This attaches the justification to the audit object. In the compact `print()` output, the corresponding checkbox is marked as completed (`[x] Independent censoring`), and `summary()` displays the full justification text beneath the assumption header.

---

## Summary

The respecified model resolved the functional form and PH issues for bilirubin, but `log(protime)` remains problematic on both counts. Several highly influential observations persist across both model specifications. In practice, the analyst would continue iterating by investigating the flagged cases, exploring alternative specifications for prothrombin time (e.g., stratification or splines), and re-auditing after each change until the remaining diagnostics are either resolved or explicitly justified.

---

## References

- Klein, J. P., & Moeschberger, M. L. (2003). *Survival Analysis: Techniques for Censored and Truncated Data* (2nd ed.). Springer.
- Therneau, T. M., & Grambsch, P. M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer.
- Xue, Y., & Schifano, E. D. (2017). Diagnostics for the Cox model. *Communications for Statistical Applications and Methods*, 24(6), 583–604.
