---
title: "Introduction to survAudit: Auditing Cox Proportional Hazards Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)
```

## Survival Analysis and the Cox Model

When conducting research where the outcome of interest is the time until an event occurs, the most widely used method is the **Cox Proportional Hazards (PH) model**. For a Cox model's estimates to be valid, the model must satisfy several structural and mathematical assumptions. 

Checking these assumptions usually requires running fragmented tests from multiple packages. `survAudit` solves this by providing:

1. **A Unified Assumption Classification**: Organizing assumptions into *Non-Identifiable* (purely qualitative), *Partially Assessable* (combining data and judgment), and *Statistically Assessable* (testable from data).
2. **Comprehensive Diagnostics**: It aggregates proportional hazards, functional form, influence, outliers, and generalized collinearity (GVIF) into one run.
3. **Auditable S3 Objects**: It returns an object containing all raw diagnostic metrics and lets analysts append qualitative justifications directly to the object before saving it.

---

## Case Study Workflow

In this vignette, we demonstrate how `survAudit` changes the analytical workflow of a typical survival analysis by revealing model limitations that are not apparent from standard Cox regression output. We use the classic `pbc` dataset from the `survival` package, which records survival data for patients with primary biliary cholangitis. Our goal is to highlight how standard Cox modeling practices can easily overlook critical assumption violations, and how the structured `survAudit` framework detects them and explicitly separates data-driven findings from those requiring scientific justification.

Importantly, `survAudit` does not just produce visual plots; it returns a rich S3 object. Analysts can directly extract the underlying data (e.g., `audit$outliers$martingale`, `audit$ph$zph`) for custom downstream analyses and to preserve a transparent diagnostic record.

### 1. Fit the Initial Model

A standard approach might begin by fitting a multivariate Cox model using a mix of continuous covariates (`age`, `bili`, `albumin`, `protime`) and a categorical factor (`edema`):

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

# Load data and convert endpoint
data(pbc, package = "survival")
pbc$status <- ifelse(pbc$status == 2, 1, 0)

# Fit a standard Cox PH model
fit <- coxph(
  Surv(time, status) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
summary(fit)
```

In many applied analyses, reporting may proceed at this stage. The standard summary output clearly indicates that age, bilirubin, albumin, and prothrombin time are all highly significant predictors of mortality (p < 0.001), along with severe edema. With interpretation focused primarily on coefficient estimates and significance tests, this model would likely be published as-is.

### 2. The survAudit Intervention

Running `survAudit()` immediately uncovers multiple violated assumptions that challenge the standard model's validity. The function aggregates goodness-of-fit, collinearity, proportional hazards, functional form, influence, and outliers into a single object.

```{r audit}
audit1 <- survAudit(fit, data = pbc)
summary(audit1)
```

#### The Ontology-Driven Summary
The summary output provides an immediate, unified view of the model's health by structuring the entire report around the **assessability ontology**. Rather than hiding the assumptions in a footnote, `survAudit` elevates them to the primary headers: Non-Identifiable, Partially Assessable, and Statistically Assessable.

For example, the summary explicitly warns that independent censoring and unmeasured confounding cannot be tested statistically and require qualitative justification based on the study design. For assumptions that are statistically or partially assessable, the summary provides the raw diagnostics (like the `cox.zph` table or GVIF) alongside immediate next steps.

#### Visual Diagnostics
Statistical tests can be overly sensitive to sample size, insensitive to localized lack of fit, or entirely absent for certain structural assumptions (such as functional form). Because of this, `survAudit`'s summary explicitly prompts the analyst to generate targeted visual diagnostics for both flagged violations and inherently visual checks.

**Overall Calibration**
First, we check the overall Goodness-of-Fit using the Cox-Snell residuals.

```{r plot-gof-initial}
plot(audit1, which = "gof")
```

The cumulative hazard of the Cox-Snell residuals generally follows the 45-degree reference line, suggesting the overall macro-level calibration of the model is acceptable, even if specific covariate assumptions are violated.

**Functional Form (Linearity)**
The summary explicitly reminds us that functional form assessment is inherently visual. If we plot the functional form using martingale residuals:

```{r plot-functional}
plot(audit1, which = "functional")
```

For `bili`, the plot clearly shows a logarithmic curve. For `protime`, the risk appears constant initially before sharply increasing. The apparent decrease at extreme values is likely an artifact of data sparsity in the right tail, highlighting why visual inspection by a domain expert is necessary.

**Proportional Hazards**
```{r plot-ph}
plot(audit1, which = "ph")
```

The Schoenfeld residual plots for both `protime` and `bili` show strong deviations from a horizontal line, making it impossible to draw a flat line within their confidence bands. This visually confirms the statistical violations flagged in the summary.

### 3. Model Respecification: An Iterative Process

Crucially, the audit provides a transparent roadmap for model respecification. First, we applied clinically plausible logarithmic transformations to bilirubin and prothrombin time to address the detected departures from linearity.

```{r respecify}
# Fit an updated model with log transformations
fit2 <- coxph(
  Surv(time, status) ~ age + log(bili) + albumin + log(protime) + factor(edema),
  data = pbc
)

# Re-audit the updated model
audit2 <- survAudit(fit2, data = pbc)
summary(audit2)
```

A subsequent re-audit of the updated model reveals the impact of these changes. Importantly, the re-audit demonstrates that model correction is iterative rather than binary: while the functional form violations were substantially reduced, the remaining proportional hazards violation for `log(protime)` indicates that additional modeling decisions—such as introducing a time-dependent effect or stratification—may still be required.

We can also verify the global goodness-of-fit of the updated model using Cox-Snell residuals:

```{r plot-gof}
plot(audit2, which = "gof")
```

### 4. The Auditable S3 Object

Importantly, `survAudit` does not just produce plots or terminal output; it returns a rich S3 object. This allows users to inspect individual diagnostic components, reuse computed quantities, and integrate audit results into custom reporting workflows.

For example, you can extract the underlying Proportional Hazards test table:

```{r extract}
print(audit2$ph$table)
```

Some assumptions (like Independent Censoring) cannot be tested statistically. `survAudit` requires you to document your qualitative justification. You can append these directly to the object:

```{r justify}
audit2$assumptions$non_identifiable$independent_censoring$justification <- 
  "Censoring was purely administrative (end of study), satisfying independent censoring."
```

If we print the audit again, this justification is now permanently attached to the object, ensuring complete transparency of both the statistics and your clinical reasoning.

By formalizing this iterative diagnostic process, `survAudit` reduces the likelihood that important model inadequacies remain undetected during routine analysis.
