---
title: "Introduction to survAudit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)
```

`survAudit` provides a structured diagnostic workflow for Cox proportional hazards models. It integrates proportional hazards testing, functional form assessment, influence and outlier diagnostics, collinearity checks, and event sufficiency into a single S3 object.

Assumptions are classified by how far they can be evaluated from the data:

- **Non-Identifiable**: Cannot be tested statistically (e.g., independent censoring). Require design-based justification.
- **Partially Assessable**: Informed by data but require analyst judgment (e.g., functional form, influence).
- **Statistically Assessable**: Evaluable via formal tests (e.g., proportional hazards, collinearity).

---

## Case Study: Primary Biliary Cholangitis

We use the `pbc` dataset from the `survival` package to illustrate a typical audit workflow.

### 1. Fit an Initial Model

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

data(pbc, package = "survival")

fit <- coxph(
  Surv(time, status == 2) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
summary(fit)
```

The model shows strong overall fit (Concordance = 0.813) and four highly significant predictors. Standard Cox output contains no information about whether model assumptions are met, it is entirely focused on estimation.

### 2. Run the Audit

Pass the fitted model and the original dataset to `survAudit()`. The `data` argument is required because partial models must be refitted when assessing functional form.

```{r audit}
audit1 <- survAudit(fit, data = pbc)
audit1
```

The compact print output provides a quick orientation. The `[ ]` checkboxes for non-identifiable assumptions are intentionally unfilled: these cannot be evaluated from data and require the analyst to supply a design-based justification (see the last section). The statistically assessable block immediately flags proportional hazards violations for `bili` and `protime`, and reports an EPV of 26.7, which is adequate.

`summary()` provides the full diagnostic report including numeric results:

```{r audit-summary}
summary(audit1)
```

The detailed report adds the full Schoenfeld test table (with per-covariate chi-squared values), the five most influential observations by max |DFBETAs|, and the five largest deviance residuals — all with the original row identifiers from `pbc`.

Working through the output section by section:

- **Data context**: `protime` has 2 missing values, so the model is fitted on 416 of 418 observations. Ten event times are tied (6.2%), which is acceptable for the Efron approximation.
- **Non-identifiable assumptions**: Both independent censoring and absence of unmeasured confounding appear as unfilled checkboxes. They require explicit justification from study design; the package cannot evaluate them.
- **Goodness-of-fit**: Only signals that the Cox-Snell residuals were computed; the visual assessment is deferred to `plot(audit, which = "gof")`.
- **Functional form**: Four continuous variables are assessed (`age`, `bili`, `albumin`, `protime`). No numeric summary is produced here — functional form is inherently visual.
- **Influence**: 69 observations exceed the $2/\sqrt{n} = 0.098$ threshold. Observation 325 has the largest maximum |DFBETAs| of 0.96, which is sizeable relative to the threshold, indicating high sensitivity of at least one coefficient to this case.
- **Outliers**: 19 observations have |deviance| > 1.96 and 9 have |log-odds| > 3.66. The largest deviance residual is 3.24 (observation 319), suggesting this patient experienced the event substantially earlier than the model predicted.
- **Proportional hazards**: The global test is strongly significant (p = 0.001). `protime` (p = 0.003) and `bili` (p = 0.005) are individually flagged. `age`, `albumin`, and `edema` show no evidence of time-varying effects.

  Two limitations of this test are worth keeping in mind: (1) it detects only linear trends in the Schoenfeld residuals over time and can miss non-monotone violations; and (2) like any chi-squared test, it becomes increasingly sensitive to minor deviations as sample size grows. Visual inspection of the residual plots is therefore essential for assessing whether a statistically flagged violation is substantively meaningful.
- **EPV**: 160 events across 6 parameters gives an EPV of 26.7, which is adequate (threshold ≥ 20).
- **Collinearity**: All VIFs are near 1, indicating no meaningful collinearity.

The most pressing issues are the proportional hazards violations and the functional form, which must be assessed visually before deciding how to address them.


### 3. Visual Diagnostics

#### Goodness of Fit

```{r plot-gof-initial}
plot(audit1, which = "gof")
```

The Cox-Snell residuals, if the model is correctly specified, should follow a unit exponential distribution. In the cumulative hazard plot, a well-fitting model will track the 45-degree reference line closely. Here the curve follows the diagonal through most of the range, indicating acceptable macro-level calibration. A modest deviation in the upper tail is common and typically reflects data sparsity at long follow-up times rather than systematic misfit. Goodness-of-fit is assessed globally, i.e. the model can pass this check and still have local violations (e.g., non-linearity in one covariate).

#### Functional Form

```{r plot-functional}
plot(audit1, which = "functional")
```

For each continuous covariate, a restricted Cox model is fitted that omits that covariate. The martingale residuals from that restricted model are then plotted against the covariate values. Because the restricted model omits the covariate, its residuals will pick up the covariate's effect on the outcome. If the relationship is correctly specified as linear, the LOESS smooth should trace an approximately straight line. A curved or non-monotone smooth indicates that the linear specification is inadequate and that a transformation or more flexible term may be needed.

For `bili`, the smooth rises sharply at low values and flattens at high values, tracing a logarithmic curve — consistent with a log-transformation. For `protime`, the smooth is approximately flat at low values before rising sharply and then dropping, again suggesting a non-linear relationship. These patterns are not apparent from the coefficient table, which assumes linearity by construction.

Common remedies for non-linearity include log-transforming the covariate, using polynomial or spline terms, or categorising the variable if a meaningful clinical threshold exists.

#### Proportional Hazards

```{r plot-ph}
plot(audit1, which = "ph")
```

The proportional hazards assumption requires that the log hazard ratio for each covariate is constant over time. The Schoenfeld residuals, after scaling, should be randomly scattered around a flat line if the assumption holds. A systematic slope or curve indicates that the hazard ratio changes over follow-up time.

The formal test (`cox.zph`) detects linear trends in the residuals over time. It will miss violations where the effect first increases and then decreases, or other non-monotone patterns. It is also sensitive to sample size: with several hundred observations, even a small and clinically irrelevant drift can yield a significant p-value. The plot is therefore the primary diagnostic; the p-value is supporting evidence.

For `bili` (p = 0.005) and `protime` (p = 0.003), the smoothed residual plots show clear trends over time. For `protime`, the hazard ratio appears to decrease over follow-up, meaning its prognostic effect is stronger early in follow-up than later. This is a substantively important finding: a single time-averaged hazard ratio obscures this time-varying behaviour. The remaining covariates show no systematic trend.


### 4. Model Respecification

The functional form plots suggest log-transformations for `bili` and `protime`. We refit and re-audit:

```{r respecify}
fit2 <- coxph(
  Surv(time, status == 2) ~ age + log(bili) + albumin + log(protime) + factor(edema),
  data = pbc
)

audit2 <- survAudit(fit2, data = pbc)
summary(audit2)
```

Comparing this report to `audit1` highlights what the transformations did and did not fix. On the positive side, the PH violation for `bili` is resolved (p = 0.306), and global collinearity remains negligible. On the negative side, the PH violation for `log(protime)` persists (p = 0.003), the global PH test is still significant (p = 0.027), and the number of flagged influential observations has increased slightly (74 vs. 69). The visual diagnostics below examine each of these in detail.

#### Functional Form Re-Assessment

```{r audit-corrected-functional}
plot(audit2, which = "functional")
```

The smooth for `log(bili)` is now approximately linear, confirming that the log scale captures the relationship adequately. The smooth for `log(protime)` still shows curvature, indicating that the log transformation did not fully resolve the non-linearity. Further exploration, such as a square-root transformation, splines, or categorisation, would be warranted before finalising the model.

#### Proportional Hazards Re-Assessment

```{r audit-corrected-ph}
plot(audit2, which = "ph")
```

The PH assumption is now satisfied for `log(bili)` (p = 0.306): the time-varying behaviour seen in the initial model was largely a consequence of the non-linear scale rather than a genuine change in the hazard ratio over time. The violation for `log(protime)` persists (p = 0.003), suggesting its effect does change over follow-up regardless of scale. Common approaches to address a genuine proportional hazards violation include stratifying the model on the offending variable or allowing its effect to vary over time by including an interaction between the covariate and a function of time (i.e., fitting a time-varying coefficient).

#### Influence

```{r influence-plot}
plot(audit2, which = "influence")
```

DFBETAs quantify, for each observation and each coefficient, how much that coefficient estimate changes when the observation is removed. Large values indicate that the estimate is sensitive to a specific case. Values exceeding the $2/\sqrt{n}$ threshold are flagged. The summary table reports the highest |DFBETAs|. The full matrix is accessible via `audit2$influence$dfbetas`. Here observations 325 and 293 have the largest maximum values.

A high DFBETAs value is not by itself a reason to exclude the observation. The appropriate response is to examine those cases individually: check for data entry errors, assess whether they represent a clinically distinct subgroup, and consider whether the estimates change substantively if they are excluded (sensitivity analysis).

#### Outliers

```{r outliers-plot}
plot(audit2, which = "outliers")
```

Outlier residuals measure how poorly the model predicts for each individual, independent of their effect on the coefficients. Deviance residuals greater than $\pm 1.96$ and log-odds residuals greater than $\pm 3.66$ are flagged. Observations with large positive deviance residuals experienced the event much sooner than the model predicted, while large negative residuals indicate the event occurred later than expected or not at all despite high predicted risk.

Several observations show deviance residuals above 3, indicating the model substantially under-predicts their risk. As with influence, these cases should be examined individually as they may reflect genuinely unusual clinical trajectories, unrecorded events, or data quality issues.

---

## Accessing the Underlying Data

All diagnostic results are stored in the returned S3 object and can be extracted directly. This is useful for custom reporting, sensitivity analyses, or integration into automated pipelines:

```{r extract}
# The full cox.zph table
audit2$ph$table

# Martingale residuals from the full model
head(audit2$outliers$martingale)

# Flagged influential observations
audit2$influence$flagged_obs
```

## Documenting Non-Identifiable Assumptions

For assumptions that cannot be tested statistically, `survAudit` provides a slot to record design-based justification directly in the object:

```{r justify}
audit2$assumptions$non_identifiable$independent_censoring$justification <-
  "Censoring was purely administrative (end-of-study cutoff). No competing events or loss to follow-up related to prognosis."
```

This attaches the justification to the audit object, making it available alongside the quantitative diagnostics when the object is saved or reported.

---

## Summary

The respecified model resolved the functional form and PH issues for bilirubin, but `log(protime)` remains problematic on both counts. Several highly influential observations persist across both model specifications. In practice, the analyst would continue iterating by investigating the flagged cases, exploring alternative specifications for prothrombin time (e.g., stratification or splines), and re-auditing after each change until the remaining diagnostics are either resolved or explicitly justified.
