---
title: "Introduction to survAudit: Auditing Cox Proportional Hazards Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)
```

## Survival Analysis and the Cox Model: A Beginner's Introduction

When conducting research where the outcome of interest is the **time until an event occurs**, standard regression methods fall short because event times are often **censored**—some subjects complete the study without ever experiencing the event.

The most widely used method for this data is the **Cox Proportional Hazards (PH) model**. Instead of predicting the exact time of an event, it estimates the **hazard ratio (HR)**. For a Cox model's hazard ratios to be valid, the model must satisfy several mathematical assumptions. If violated, conclusions can be severely biased.

---

## Why Use `survAudit`?

Checking Cox model assumptions usually requires running fragmented tests from multiple packages. `survAudit` solves this by implementing:

1. **A Unified Assumption Classification**: Organizing assumptions into *Non-Identifiable* (purely qualitative), *Partially Assessable* (combining data and judgment), and *Statistically Assessable* (testable from data).
2. **Comprehensive Diagnostics**: It aggregates proportional hazards, functional form, influence, outliers, and generalized collinearity (GVIF) into one run.
3. **Auditable R Objects**: It returns an S3 object containing all raw diagnostic metrics (allowing custom extraction) and lets analysts append qualitative justifications directly to the object before saving it.

---

## Step-by-Step Tutorial

We will demonstrate `survAudit` using the classic `pbc` dataset from the `survival` package, which records survival data for patients with primary biliary cholangitis.

### 1. Fit the Cox Model
We fit a Cox model using a mix of continuous covariates (`age`, `bili`, `albumin`, `protime`) and a categorical factor (`edema`):

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

# Load data and convert endpoint
data(pbc, package = "survival")
pbc$status <- ifelse(pbc$status == 2, 1, 0)

# Fit a naive Cox PH model
fit <- coxph(
  Surv(time, status) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
```

### 2. Run the Audit
Run the diagnostic audit using `survAudit()`.

> [!NOTE]
> Supplying `data = pbc` is optional but highly recommended to ensure complete case extraction for functional form plots.

```{r audit}
audit <- survAudit(fit, data = pbc)
```

### 3. Review the Summary
The `summary()` method prints a concise console dashboard detailing the status of every assumption:

```{r summary}
summary(audit)
```

### 4. Visual Diagnostics
While the summary provides immediate flags, `survAudit` provides a dedicated `plot()` method to visually explore the violations. 

The `which` argument controls what is displayed:
- `"gof"`: Cox-Snell residuals for overall calibration.
- `"ph"`: Scaled Schoenfeld residuals.
- `"functional"`: Martingale residuals with a LOESS smooth for checking linearity.
- `"influence"`: DFBETAs and Likelihood Displacement.
- `"outliers"`: Deviance residuals.

For example, to visually inspect the functional form of our continuous covariates:

```{r plot-functional}
plot(audit, which = "functional")
```

The plots for `bili` and `protime` clearly show a logarithmic curve rather than a straight line, indicating the model is mis-specified and needs log transformations.

### 5. Extracting Raw Data for Custom Analysis
`survAudit` is not a black-box plotter. The `audit` object is a rich list containing all the underlying metrics. You can extract these directly to build custom plots or tables:

```{r extract}
# Extract the underlying Proportional Hazards test table
print(audit$ph$table)

# Extract the deviance residuals to find outliers
deviance_res <- audit$outliers$deviance
max_outlier <- which.max(abs(deviance_res))
print(paste("Largest outlier is at index:", max_outlier))
```

### 6. Appending Qualitative Justifications
Some assumptions (like Independent Censoring) cannot be tested statistically. `survAudit` requires you to document your qualitative justification.

You can modify the object directly in R:

```{r justify}
audit$assumptions$non_identifiable$independent_censoring$justification <- 
  "Censoring was purely administrative (end of study), satisfying independent censoring."
```

If we print the audit again, this justification is now permanently attached to the object:

```{r print-justification}
print(audit)
```

You can now save this object (e.g., `saveRDS(audit, "model_audit.rds")`) and pass it to reviewers, ensuring complete transparency of both the statistics and your clinical reasoning.
