---
title: "Introduction to survAudit: Auditing Cox Proportional Hazards Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to survAudit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE
)
```

## Survival Analysis and the Cox Model

When conducting research where the outcome of interest is the time until an event occurs, the most widely used method is the **Cox Proportional Hazards (PH) model**. For a Cox model's estimates to be valid, the model must satisfy several structural and mathematical assumptions. 

Checking these assumptions usually requires running fragmented tests from multiple packages. `survAudit` solves this by providing:

1. **A Unified Assumption Classification**: Organizing assumptions into *Non-Identifiable* (purely qualitative), *Partially Assessable* (combining data and judgment), and *Statistically Assessable* (testable from data).
2. **Comprehensive Diagnostics**: It aggregates proportional hazards, functional form, influence, outliers, and generalized collinearity (GVIF) into one run.
3. **Auditable S3 Objects**: It returns an object containing all raw diagnostic metrics and lets analysts append qualitative justifications directly to the object before saving it.

---

## Case Study Workflow

In this vignette, we demonstrate how `survAudit` changes the analytical workflow of a typical survival analysis by revealing model limitations that are not apparent from standard Cox regression output. We use the classic `pbc` dataset from the `survival` package, which records survival data for patients with primary biliary cholangitis. Our goal is to highlight how standard Cox modeling practices can easily overlook critical assumption violations, and how the structured `survAudit` framework detects them and explicitly separates data-driven findings from those requiring scientific justification.

Importantly, `survAudit` does not just produce visual plots; it returns a rich S3 object. Analysts can directly extract the underlying data (e.g., `audit$outliers$martingale`, `audit$ph$zph`) for custom downstream analyses and to preserve a transparent diagnostic record.

### 1. Fit the Initial Model

A standard approach might begin by fitting a multivariate Cox model using a mix of continuous covariates (`age`, `bili`, `albumin`, `protime`) and a categorical factor (`edema`):

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(survAudit)

# Load data and convert endpoint
data(pbc, package = "survival")
pbc$status <- ifelse(pbc$status == 2, 1, 0)

# Fit a standard Cox PH model
fit <- coxph(
  Surv(time, status) ~ age + bili + albumin + protime + factor(edema),
  data = pbc
)
summary(fit)
```

In many applied analyses, reporting may proceed at this stage. The standard summary output clearly indicates that age, bilirubin, albumin, and prothrombin time are all highly significant predictors of mortality (p < 0.001), along with severe edema. With interpretation focused primarily on coefficient estimates and significance tests, this model would likely be published as-is.

### 2. The survAudit Intervention

Running `survAudit()` immediately uncovers multiple violated assumptions that challenge the standard model's validity. The function aggregates goodness-of-fit, collinearity, proportional hazards, functional form, influence, and outliers into a single object.

```{r audit}
audit1 <- survAudit(fit, data = pbc)
summary(audit1)
```

#### Visual Diagnostics
Statistical tests can be overly sensitive to sample size, insensitive to localized lack of fit, or entirely absent for certain structural assumptions (such as functional form). Because of this, `survAudit`'s summary explicitly prompts the analyst to generate targeted visual diagnostics for both flagged violations and inherently visual checks.

**Overall Calibration**
First, we check the overall Goodness-of-Fit using the Cox-Snell residuals.

```{r plot-gof-initial}
plot(audit1, which = "gof")
```

The cumulative hazard of the Cox-Snell residuals generally follows the 45-degree reference line, suggesting the overall macro-level calibration of the model is acceptable, even if specific covariate assumptions are violated.

**Functional Form (Linearity)**
The summary explicitly reminds us that functional form assessment is inherently visual. If we plot the functional form using martingale residuals:

```{r plot-functional}
plot(audit1, which = "functional")
```

For `bili`, the plot clearly shows a logarithmic curve. For `protime`, the risk appears constant initially before sharply increasing. The apparent decrease at extreme values is likely an artifact of data sparsity in the right tail, highlighting why visual inspection by a domain expert is necessary.

**Proportional Hazards**
```{r plot-ph}
plot(audit1, which = "ph")
```

The Schoenfeld residual plots for both `protime` and `bili` show strong deviations from a horizontal line. This visually confirms the statistical violations flagged in the summary.

**Summary of Initial Audit**
Without having seen all the plots already, the initial `survAudit` explicitly revealed that the standard Cox model suffers from multiple structural issues: 

1. Non-linearity for both `bili` and `protime`.
2. Proportional hazards violations for `bili` and `protime`.
3. The presence of several highly influential observations and outliers across the dataset.

Rather than proceeding with the initial model, the analyst is now equipped with the exact information needed to iteratively correct the model. As a first step, we will apply log-transformations to `bili` and `protime` to address the functional form violations, before we re-audit the model. 

### 3. Iterative Model Correction

Guided by the audit, we fit a corrected model using the log-transformed variables:

```{r respecify}
# Fit an updated model with log transformations
fit2 <- coxph(
  Surv(time, status) ~ age + log(bili) + albumin + log(protime) + factor(edema),
  data = pbc
)

# Re-audit the updated model
audit2 <- survAudit(fit2, data = pbc)
summary(audit2)
```

The initial audit suggested a log-linear relationship for bilirubin, but we must re-check all diagnostic plots to confirm the transformation's effect across all assumptions.

**Overall Calibration Re-Assessment**
```{r audit-corrected-gof}
plot(audit2, which = "gof")
```
The overall goodness-of-fit using Cox-Snell residuals continues to follow the 45-degree reference line except for very few outliers, confirming that the macro-level calibration of the model remains acceptable after the variable transformations.

**Functional Form Re-Assessment**
```{r audit-corrected-functional}
plot(audit2, which = "functional")
```
The functional form plots for the updated model show that the log-transformations were not a complete fix. While the curve for `log(bili)` shows some improvement, the functional form for `log(protime)` remains problematic. (Note: Ensure you cycle through or extract all generated plot panels to view every continuous covariate).

**Proportional Hazards Re-Assessment**
```{r audit-corrected-ph}
plot(audit2, which = "ph")
```
Re-assessing proportional hazards reveals further complications. The proportional hazards assumption is still not satisfied for `log(bili)`. Interestingly, the Schoenfeld residual slope for `log(protime)` looks a tiny bit better, but the assumption is still violated. 

**Influence Re-Assessment**
```{r influence-plot}
plot(audit2, which = "influence")
```
When evaluating influence, analysts should visually inspect the DFBETAs plots for points that stray far from the main cluster and cross the dotted threshold lines. The audit explicitly reports that multiple observations still exceed the conservative DFBETAs threshold. 

**Outliers Re-Assessment**
```{r outliers-plot}
plot(audit2, which = "outliers")
```
For outliers, the audit generates a four-panel plot displaying Deviance, Log-Odds, Martingale, and Normal Deviate residuals against the linear predictor. Analysts should check these residual plots for points that systematically exceed their respective threshold bands (e.g., $\pm 1.96$ for deviance and normal deviate, or $\pm 3.66$ for log-odds). Several observations continue to exhibit extreme deviance residuals. 

These findings emphasize that influence and outliers are global properties of the data. It is highly recommended that these extreme outliers and influential points are checked manually by the analyst to ensure they are not the result of data entry errors, and to evaluate their impact on the final estimates for example by using sensitivity analyses.

**Ongoing Model Refinement**
The initial analysis masked severe misspecifications, but `survAudit` surfaced critical violations. Our re-audit demonstrated that survival modeling is highly iterative—simple log-transformations are rarely a silver bullet. While some assumptions improved, proportional hazards violations and influential outliers persisted.

By formalizing these checks, `survAudit` prevents the premature publication of mis-specified models and guarantees transparency. We are not finished here: the logical next steps involve manually investigating the highly influential observations and exploring alternative strategies for both the `bili` and `protime` variables (e.g., stratification or time-dependent effects) to achieve a robust final model.

### 4. The Auditable S3 Object

Importantly, `survAudit` does not just produce plots or terminal output; it returns a rich S3 object. This allows users to inspect individual diagnostic components, reuse computed quantities, and integrate audit results into custom reporting workflows.

For example, you can extract the underlying Proportional Hazards test table:

```{r extract}
print(audit2$ph$table)
```

Some assumptions (like Independent Censoring) cannot be tested statistically. `survAudit` requires you to document your qualitative justification. You can append these directly to the object:

```{r justify}
audit2$assumptions$non_identifiable$independent_censoring$justification <- 
  "Censoring was purely administrative (end of study), satisfying independent censoring."
```

If we print the audit again, this justification is now permanently attached to the object, ensuring complete transparency of both the statistics and your clinical reasoning.

By formalizing this iterative diagnostic process, `survAudit` reduces the likelihood that important model inadequacies remain undetected during routine analysis.
